Simple repression draft

### 2020-05-21 doc/section_04_bayesian_inference.tex
MRM pass through section 4 on 20200521

% !TEX root = ./busty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model selection}
\subsection{Parameter inference for constitutive promoters}

From consideration of Fano factors in the previous section, we suspect that
model 5 in~\fig{fig:constit_cartoons}, a one-state bursty model of constitutive
promoters, achieves a Goldilocks level of complexity. Does this stand up to
closer scrutiny, namely, comparison to full mRNA distributions rather than
simply their moments? We will test this thoroughly on all the data from
constitutive promoters from Jones et.\ al.~\cite{Jones2014}.

It will be instructive, however, to first consider the Poisson promoter, model 1
in~\fig{fig:constit_cartoons}. As we alluded to earlier, since the Poisson
distribution has its Fano factor $\nu$ strictly equal to 1,
and all of the observed data has Fano factor $\nu>1$, we might
already suspect that this model is incapable of fitting the data.
We will verify that this in fact the case. As a consequence
of ruling out model 1, models 2 and 3 will also be ruled out.
Why? Again all the data have $\nu>1$, meaning the mRNA distributions
are overdisperse compared with a Poisson distribution,
while models 2 and 3 have Fano factors $\nu\le 1$ meaning they
are underdisperse relative to Poisson. So models 2 and 3
can fare no better than the simple Poisson model. We will also not explicitly
consider model 4 from~\fig{fig:constit_cartoons} since it was already thoroughly
analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
case of it.

In performing parameter inference on the FISH data of~\cite{Jones2014}, building
a generative model is straightforward. The data are single-cell mRNA counts, and
we assume different cells to be independent and identically distributed from the
underlying generative model, which is just the steady-state mRNA distribution
computed from the master equation of the underlying model.

\subsubsection{Model 1: Poisson promoter}
\mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
For this model the master equation of interest is~\eq{eq:poisson_promoter_cme}
with repressor set to zero, i.e.,
\begin{equation}
\deriv{t}p_{m,U}(t) = rp_{m-1,U}(t) - rp_{m,U}(t)
        + (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
\end{equation}
whose steady-state distribution is Poisson, given by
\begin{equation}
p_U(m) = \frac{\lambda^m e^{-\lambda}}{m!}
\end{equation}
with $\lambda=r/\gamma$. This steady state distribution defines our likelihood
$p(m\mid\lambda)$, i.e.
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}
\label{eq:poisson_inference010}
\end{equation}
is the probability of observing a single cell with $m$ mRNAs given a known value
of $\lambda$, written to make the conditioning explicit.

With our likelihood in hand, we can invoke Bayes theorem to write down the
posterior distribution on our model parameter $\lambda$, given by
\begin{equation}
p(\lambda\mid D) \propto p(D\mid\lambda) p(\lambda),
\end{equation}
where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts over a
population of $N$ identical cells. As is standard we have neglected the factor
$p(D)$ on the right hand side since it is independent of $\lambda$ and serves
only as a normalization factor. Since we assume each cell's mRNA count is
independent of others, the likelihood is simply a product of the single cell
likelihoods given by \eq{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}
This is often notated simply as
\begin{equation}
D \sim \text{Poisson}(\lambda),
\end{equation}
which is read as the data $D$ is Poisson
distributed with mean $\lambda$. The $\sim$ operator literally denotes ``is
distributed according to.''
The conditioning on $\lambda$ is no longer explictily notated,
only implied by its appearance on the right hand side.

To proceed we need to specify a prior. In this case we are extremely data-rich,
as the dataset from Jones et.\ al~\cite{Jones2014} has of order 1000-3000
single-cell measurements for each promoter, so our choice of prior matters
little here, as long as it does not exclude the true value of $\lambda$. A
convenient choice for our problem is to use a \textit{conjugate} prior.
A conjugate prior is a special prior that causes the posterior to
have the same functional form as the prior, simply with updated
model parameters. This makes calculations
analytically tractable and also offers a nice interpretation of the inference
procedure as updating our knowledge about the model parameters.
This makes conjugate priors very useful when they exist.
The caveat is that conjugate priors only exist for a very limited
number of likelihoods, mostly with only one or two model
parameters, so in almost all other Bayesian inference problems,
we must tackle the posterior numerically.

But, for the problem at hand, a conjugate prior does in fact exist.
For a Poisson likelihood of identical and identically distributed
data, the conjugate prior is a gamma distribution, as
can be looked up in, e.g.,~\cite{Gelman2013}, Section 2.6.
Putting a gamma prior on $\lambda$ introduces two new parameters $\alpha$ and
$\beta$ which parametrize the gamma distribution itself, which we use to encode
the range of $\lambda$ values we view as reasonable. Recall $\lambda$ is the
mean steady-state mRNA count per cell, which \textit{a priori} could plausibly
be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$ achieve this,
since the gamma distribution is strictly positive with mean $\alpha/\beta$ and
standard deviation $\sqrt{\alpha}/\beta$. To be explicit, then, our prior is
\begin{equation}
\lambda \sim \text{Gamma}(\alpha, \beta)
\end{equation}

As an aside, note that if we did not know that our prior was a conjugate prior,
we could still write down our posterior distribution from its definition as
\begin{equation}
p(\lambda\mid D,\alpha,\beta)
\propto p(D\mid\lambda) p(\lambda \mid\alpha,\beta)
\propto \left(\prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}\right)
        \frac{\beta}{\Gamma(\alpha)}(\beta\lambda)^{\alpha-1} e^{-\beta\lambda}
.
\end{equation}
Without foreknowledge that this in fact reduces to a gamma distribution, this
expression might appear rather inscrutable. When conjugate priors are
unavailable for the likelihood of interest - which is almost always the case for
models with $>1$ model parameter - this inscrutability is the norm, and making
sense of posteriors analytically is almost always impossible. Fortunately, MCMC
sampling provides us a powerful method of constructing posteriors numerically
which we will make use of extensively.

Since we did use a conjugate prior, we may simply look up our posterior in any
standard reference such as~\cite{Gelman2013}, Section 2.6,
from which we find that
\begin{equation}
\lambda
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
\end{equation}
where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$ for
notational convenience. A glance at the FISH data from~\cite{Jones2014} reveals
that $N$ is $\mathcal{O}(10^3)$ and $\langle m\rangle \gtrsim 0.1$ for all
constitutive strains in~\cite{Jones2014}, so $\bar{m}N \gtrsim 10^2$. Therefore
as we suspected, our prior parameters are completely overwhelmed by the data.
The prior behaves, in a sense, like $\beta$ extra ``data points''
with a mean value of $(\alpha-1)/\beta$~\cite{Gelman2013}, which
gives us some intuition for how much data is needed to overwhelm
the prior in this case: enough data $N$ such that $\beta\ll N$
and $\alpha/\beta \ll \bar{m}$. In
fact, $\bar{m}N$ and $N$ are so large that we can, to an excellent
approximation, ignore the $\alpha$ and $\beta$ dependence and approximate the
gamma distribution as a Gaussian with mean $\bar{m}$ and standard deviation
$\sqrt{\bar{m}/N}$, giving
\begin{equation}
\lambda
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
\approx \text{Normal}\left(\bar{m}, \sqrt{\frac{\bar{m}}{N}}\right).
\end{equation}
As an example with real numbers, for the \textit{lacUV5} promoter, Jones et.\
al~\cite{Jones2014} measured 2648 cells with an average mRNA count per cell of
$\bar{m} \approx 18.7$. In this case then, our posterior is
\begin{equation}
\lambda
\sim \text{Normal}\left(18.7, 0.08\right),
\end{equation}
which suggests we have inferred our model's one parameter to a precision of
order 1\%.

This is not wrong, but it is not the full story. The model's posterior is
tightly constrained, but is it a good generative model? In other words, does the
model generate data that look similar to our actual data, and is it therefore
plausible that the model captures the important features of the data generating
process? This intuitive notion can be codified with \textit{posterior predictive
checks}, or PPCs, and we will see that this simple Poisson model fails badly.

The intuitive idea of posterior predictive checks is simple: \mmnote{Not sure if
all this stuff is clearer by example, but then does it confuse the reader in
generalizing to other models?}
\begin{enumerate}
\item Make a random draw of the model parameter $\lambda$ from the posterior
distribution.
\item Plug that draw into the likelihood and generate a synthetic dataset
$\{m_k\}$ conditioned on $\lambda$.
\item Repeat many times.
\end{enumerate}
More formally, the posterior predictive distribution can be thought of as the
distribution of future yet-to-be-observed data, conditioned on the data we have
already observed. Clearly if those data appear quite different, the model has a
problem. Put another way, if we suppose the generative model is true, then the
synthetic datasets we generate should resemble the actual observed data, and if
not, it suggests the model is missing important features. All the data we
consider in this work are 1D (distributions of mRNA counts over a population) so
ECDFs are an excellent visual means of comparing synthetic and observed
datasets. In general for higher dimensional datasets, much of the challenge is
in merely designing good visualizations that can actually show if synthetic and
observed data are similar or not.

For our example Poisson promoter model then, we merely draw many random numbers,
say 1000, from the Gaussian posterior. For each one of those draws, we generate
a dataset from the likelihood, i.e., we draw 2648 (the number of observed cells
in the actual dataset) Poisson-distributed numbers for each of the 1000
posterior draws, for a total of 2648000 samples from the posterior predictive
distribution.

To compare so many samples with the actual observed data, one excellent
visualization for 1D data is ECDFs of the quantiles, as shown for our Poisson
model in~\fig{fig:constit_post_full}.
In this example, the median for each possible mRNA
count is shown as a dark green line, while shaded bands of lighter
green contain 95\% of the posterior predictive samples.
Plotting quantiles in this way gives us a sense of the range of data we might
consider plausible, under the assumption that the model is true. In this case it
is quite obvious that the observed data, plotted in orange, could not plausibly
come from this Poisson generative model. The other model with a PPC plotted
in~\fig{fig:constit_post_full}, bursty model 5 with
a negative binomial steady-state mRNA
distribution, looks like a good candidate by eye. We cover it in more detail
later.

\subsubsection{Nonequilibrium Model Five - Bursty promoter}
Let us now consider the problem of parameter inference from FISH data
for model five from~\fig{fig1:means_cartoons}(C). As derived
in Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution
in this model is a negative binomial distribution,
given by~\eq{eq:nbinom_deriv_final}, which is
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate
nondimensionalized by the mRNA degradation rate $\gamma$.
As sketched earlier, the story of this distribution is of
geometrically-distributed bursts of mRNA, where the arrival of bursts is a
Poisson process with rate $k_i$ and the mean size of a burst is $b$.

As for the Poisson promoter model, this expression for the steady-state
mRNA distribution is exactly the likelihood we want to use in Bayes theorem.
Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$,
here Bayes theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood so we only need to choose priors
on $k_i$ and $b$. For the datasets from~\cite{Jones2014} that we
are analyzing, as for the Poisson promoter model above we are
still data-rich so the prior's influence remains weak, but not
nearly as weak because the dimensionality of our model has
increased from one to two.

We follow the guidance of~\cite{Gelman2013}, Section 2.9 in
opting for weakly-informative priors on $k_i$ and $b$ (conjugate
priors do not exist for this problem), and we find
``street-fighting estimates''~\cite{Mahajan2010} to be an ideal
way of constructing such priors. The idea of weakly informative
priors is to allow all remotely plausible values of model
parameters while excluding the completely absurd or unphysical.

Consider $k_i$. The fastest known bacterial promoters control
rRNA genes and initiate transcripts no faster than $\sim
1/\text{sec}$. It would be exceedingly strange if any of the
constitutive promoters from~\cite{Jones2014} were stronger than
that, so we can take that as an upper bound. For a lower bound,
if transcripts are produced too rarely, there would be nothing to
see with FISH. The datasets for each strain contain of order
$10^3$ cells, and if the $\langle m \rangle = k_i b/\gamma
\lesssim 10^{-2}$, then the total number of expected mRNA
detections would be single-digits or less and we would have
essentially no data on which to carry out inference. So assuming
$b$ is not too different from 1, justified next, and an mRNA
lifetime of $\gamma^{-1}\sim 3-5~\text{min}$, this gives us soft
bounds on $k_i/\gamma$ of perhaps $10^{-2}$ and $3\times 10^1$.

Next consider mean burst size $b$. This parametrization of the
geometric distribution allows bursts of size zero (which could
representing aborted transcripts and initiations), but it would
be quite strange for the mean burst size $b$ to be below
$\sim10^{-1}$, for which nearly all bursts would be of size zero
or one. For an upper bound, if transcripts are initiating at a
rate somewhat slower than rRNA promoters, then it would probably
take a time comparable to the lifetime of an mRNA to produce a
burst larger than 10-20 transcripts, which would invalidate the
approximation of the model that the duration of bursts are
instantaneous compared to other timescales in the problem. So we
will take soft bounds of $10^{-1}$ and $10^1$ for $b$.

Note that the natural scale for these ``street-fighting estimates''
was a log scale. This is commonly the case that our prior sense
of reasonable and unreasonable parameters is set on a log scale.
A natural way to enforce these soft bounds is therefore to use a
lognormal prior distribution, with the soft bounds set $\pm2$
standard deviations from the mean.

With this, we are ready to write our full generative model as
\begin{equation}
\begin{split}
\ln k_i \sim \text{Normal}(-0.5, 2)
\\
\ln b \sim \text{Normal}(0.5, 1)
\\
m \sim \text{NBinom}(k_i, b).
\end{split}
\end{equation}
We carried out MCMC sampling on the posterior of this model, starting
with the constitutive \textit{lacUV5} dataset from~\cite{Jones2014}.
The resulting MCMC samples are shown in~\fig{fig:constit_post_full}(A).
In constrast to the active/inactive constitutive model considered
in~\cite{Razo-Mejia2020} (nonequilibrium model four
in~\ref{fig:constit_cartoons}), this model is well-identified
with both parameters determined to a fractional uncertainty of 5-10\%.
The strong correlation reflects the fact that their product sets
the mean of the mRNA distribution, which is tightly constrained
by the data, but there is weak ``sloppiness''~\cite{Transtrum2015}
along a set of values with a similar product.

Flush with success having found the model's posterior to be
well-identified, the next step was posterior predictive sampling.
The only procedural difference compared to the Poisson promoter
model is that, since we have no analytical posterior expression
but only samples from the posterior, we generate a synthetic
dataset (i.e., posterior predictive samples) for each posterior
sample from the MCMC sampling. \fig{fig:constit_post_full}(B)
shows the resulting predictive ECDF; similarly to the Poisson
promoter model, the median posterior predictive ECDF is plotted
as a dark blue line, while a lighter blue shaded region
encompasses 95\% of the posterior predictive samples. Unlike the
Poisson promoter model, the experimental ECDF closely tracks the
posterior predictive ECDF, indicating this model is actually able
to generate the observed data and increasing our confidence that
this model is at least not wrong.

\textit{lacUV5} is our primary target here, since it forms the
core of all the simple repression constructs of~\cite{Jones2014}
that we consider in Section~\ref{sec:rep_kinetics_inference}.
Nevertheless, we thought it wise to apply our bursty promoter
model to the other 17 constitutive promoters available in the
FISH dataset from~\cite{Jones2014} as a test that the model is
capturing the essential phenomenology. If the model fit well to
all the different promoters, this would increase our confidence
that it would serve well as a foundation for inferring repressor
kinetics later in Section~\ref{sec:rep_kinetics_inference}.
Conversely, were the model to fail on more than a couple of the
other promoters, it would give us pause.

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
        comparison.}
        (A) The joint posterior density of model 5, the
        bursty promoter with negative binomially-distributed
        steady state, is plotted with MCMC samples.
        1D marginal densities are plotted as flanking histograms.
        The model was fit on \textit{lacUV5} data from~\cite{Jones2014}.
        (B) The ECDF of the observed population distribution of
        mRNA transcripts under the control of a constitutive
        lacUV5 promoter is shown in orange.
        The median posterior predictive ECDFs for models (1),
        Poisson, and (5), negative binomial, are plotted in
        green and blue, respectively.
        Lighter green and blue regions enclose 95\% of all
        posterior predictive samples from their respective models.
        Model (1) is in obvious contradiction with the data
        while model (5) is not.
        FISH data is again from~\cite{Jones2014}.
    (C) Joint posterior distributions for burst rate $k_i$ and mean burst size
    $b$ for 18 promoters from~\cite{Jones2014}. Each contour indicates the 95\%
    highest posterior density region for a particular promoter. Note that the
    vertical axis is shared with (D). Panel (D) plots the burst rate $k_i$ vs.\
    the binding energy for each promoter from energy matrices
    from~\cite{Brewster2012}.
    The dotted line shows the predicted slope according to
    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
    described in text.}
\label{fig:constit_post_full}
\end{figure}

>>>>

% !TEX root = ./bursty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model 
selection}
\subsection{Parameter inference for constitutive promoters}

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig:constit_cartoons}, a one-state bursty model of
constitutive promoters, achieves a Goldilocks level of complexity. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on
single-cell mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
the observed data has Fano factor $\nu>1$, we might already suspect that this
model is incapable of fitting the data. We will verify that this in fact the
case. Using the same argument we can immediately rule out models 2 and 3 from 
Figure~\ref{fig:constit_cartoons}. these models have Fano factors $\nu\le 1$
meaning they are underdisperse relative to Poisson. We will also not explicitly
consider model 4 from~\fig{fig:constit_cartoons} since it was already thoroughly
analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
case of it.

In performing parameter inference on the single-molecule data
of~\cite{Jones2014}, building a generative model is straightforward. The data
are single-cell mRNA counts, and we assume different cells to be independent and
identically distributed from the underlying generative model. The generative
model for our case is the steady-state mRNA distribution computed from the
master equation under consideration.

\subsubsection{Model 1: Poisson promoter}
% \mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
For this model the master equation of interest is
Eq.~\ref{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
\begin{equation}
\deriv{t}p_{m,U}(t) = rp_{m-1,U}(t) - rp_{m,U}(t)
        + (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
\end{equation}
whose steady-state solution is given by a Poisson distribution with parameter
$\lambda \equiv r / \gamma$. This steady state distribution defines the
likelihood $p(m\mid\lambda)$ for our Bayesian inference, i.e. 
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}
\label{eq:poisson_inference010}
\end{equation}
is the probability of observing a single cell with $m$ mRNAs given a value of
$\lambda$, written to make the conditioning explicit.

With our likelihood in hand, we can invoke Bayes theorem to write down the
posterior distribution on our model parameter $\lambda$, given by
\begin{equation}
p(\lambda\mid D) \propto p(D\mid\lambda) p(\lambda),
\end{equation}
where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts over a
population of $N$ identical cells. As is standard we have neglected the factor
$p(D)$ on the right hand side since it is independent of $\lambda$ and serves
only as a normalization factor. Since we assume each cell's mRNA count is
independent of others, the likelihood of the full inference problem
$p(D\mid\lambda)$ is simply a product of the single cell likelihoods given by
\eq{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}
This is often notated simply as
\begin{equation}
D \sim \text{Poisson}(\lambda),
\end{equation}
which is read as the data $D$ is Poisson distributed with mean $\lambda$. The
$\sim$ operator literally denotes ``is distributed according to.'' The
conditioning on $\lambda$ is no longer explicitly notated, only implied by its
appearance on the right hand side.

To proceed we need to specify a prior. In this case we are extremely data-rich,
as the dataset from Jones et.\ al~\cite{Jones2014} has of order 1000-3000
single-cell measurements for each promoter, so our choice of prior matters
little here, as long as it is sufficiently broad. For details on the prior 
selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
purpose here it suffices to specify that we use a prior of the form $\lambda
\sim \text{Gamma}(\alpha, \beta)$. Putting a gamma prior on $\lambda$ introduces
two new parameters $\alpha$ and $\beta$ which parametrize the gamma distribution
itself, which we use to encode the range of $\lambda$ values we view as
reasonable. Recall $\lambda$ is the mean steady-state mRNA count per cell, which
\textit{a priori} could plausibly be anywhere from 0 to a few hundred.
$\alpha=1$ and $\beta=1/50$ achieve this, since the gamma distribution is
strictly positive with mean $\alpha/\beta$ and standard deviation
$\sqrt{\alpha}/\beta$.

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior
known as the \textit{conjugate} prior for a Poisson likelihood allows us to 
get a closed form for the posterior distribution $p(\lambda \mid D)$. In 
particular it can be shown that~\cite{Gelman2013}
\begin{equation}
\lambda
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
\end{equation}
where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$ for
notational convenience. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to make the 
excellent approximation 
\begin{equation}
\lambda
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
\approx \text{Normal}\left(\bar{m}, \sqrt{\frac{\bar{m}}{N}}\right).
\end{equation}
As an example with real numbers, for the \textit{lacUV5} promoter, Jones et.\
al~\cite{Jones2014} measured 2648 cells with an average mRNA count per cell of
$\bar{m} \approx 18.7$. In this case then, our posterior is
\begin{equation}
\lambda
\sim \text{Normal}\left(18.7, 0.08\right),
\label{eq:posterior_poisson}
\end{equation}
which suggests we have inferred our model's one parameter to a precision of
order 1\%.

\mrm{
We alert the reader not to confuse a tightly constrained posterior distribution
with a good generative model. In other words, not because the single parameter
for our Poisson likelihood can only take a small range of values constrained by
the data, this implies that such Poisson distribution characterizes the
underlying cell-to-cell variability we observe in the data. To systematically
check the validity of the generative model we can make use of \textit{posterior
predictive checks}. These consists on generating synthetic data using our
generative model for many plausible values of our parameter. For the case of our
\textit{lacUV5} promoter this means that we sample a value for $\lambda$ out of
the posterior distribution (Eq.~\ref{eq:posterior_poisson}), and generate 2648
samples out of a Poisson distribution with such $\lambda$ value. We then repeat
this procedure 1000 times. Clearly if those data appear quite different, the
model has a problem. Put another way, if we suppose the generative model is
true, then the synthetic datasets we generate should resemble the actual
observed data, and if not, it suggests the model is missing important features.}

\mrm{
To compare so many samples with the actual observed data, we can use empirical
cumulative distribution functions (ECDF) of the distribution quantiles. This 
representation is shown in Figure~\ref{fig:constit_post_full}(B). In this 
example, the median for each possible mRNA count for the Poisson distribution
is shown as a dark green line, while the lighter green contains 95\% of the
posterior predictive samples. This way of representing the posterior predictive
checks gives us a sense of the range of data we might consider plausible, under
the assumption that the model is true. In this case it is quite obvious that the
observed data, plotted in orange, could not plausibly come from this Poisson
generative model. We will cover the other model shown in
Figure~\ref{fig:constit_post_full}(B) next.
}

\subsubsection{Model 5 - Bursty promoter}
Let us now consider the problem of parameter inference from FISH data for model
five from~\fig{fig1:means_cartoons}(C). As derived in
Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
model is a negative binomial distribution, given by
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\label{eq:neg_bionom}
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate nondimensionalized
by the mRNA degradation rate $\gamma$. As sketched earlier, the story of this
distribution is of geometrically-distributed bursts of mRNA, where the arrival
of bursts is a Poisson process with rate $k_i$ and the mean size of a burst is
$b$.

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use in Bayes theorem. Again
denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
Bayes theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood --the product of $N$ negative binomials as
Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
promoter model above we are still data-rich so the prior's influence remains
weak, but not nearly as weak because the dimensionality of our model has
increased from one to two.
\mrm{
Details on the arguments behind our prior distribution selection are left for
Appendix~\ref{sec:bayesian}. Our full generative model can be specified as
\begin{equation}
\begin{split}
\ln k_i \sim \text{Normal}(-0.5, 2) \\
\ln b \sim \text{Normal}(0.5, 1) \\
m \sim \text{NBinom}(k_i, b).
\end{split}
\end{equation}
Note that we specify a distribution over the logarithm of our parameters $k_i$
and $b$. This is commonly the case for parameters for which our previous 
knowledge based on our domain expertise spans several orders of magnitude.
}

We carried out MCMC sampling on the posterior of this model, starting with the
constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
samples are shown in~\fig{fig:constit_post_full}(A). In contrast to the
active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
(nonequilibrium model 4 in~\ref{fig:constit_cartoons}), this model is
well-identified with both parameters determined to a fractional uncertainty of
5-10\%. The strong correlation reflects the fact that their product sets the
mean of the mRNA distribution, which is tightly constrained by the data, but
there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
similar product.

Flush with success having found the model's posterior to be well-identified, the
next step is posterior predictive sampling. The only procedural difference
compared to the Poisson promoter model is that, since we have no analytical
posterior expression but only samples from the posterior, we generate a
synthetic dataset (i.e., posterior predictive samples) for each posterior sample
from the MCMC sampling. \fig{fig:constit_post_full}(B) shows the resulting
predictive ECDF; similarly to the Poisson promoter model, the median posterior
predictive ECDF is plotted as a dark blue line, while a lighter blue shaded
region encompasses 95\% of the posterior predictive samples. Unlike the Poisson
promoter model, the experimental ECDF closely tracks the posterior predictive
ECDF, indicating this model is actually able to generate the observed data and
increasing our confidence that this model is at least not wrong.

\textit{lacUV5} is our primary target here, since it forms the core of all the
simple repression constructs of~\cite{Jones2014} that we consider in
Section~\ref{sec:rep_kinetics_inference}. Nevertheless, we thought it wise to
apply our bursty promoter model to the other 17 unregulated promoters available
in the single-cell mRNA count dataset from~\cite{Jones2014} as a test that the
model is capturing the essential phenomenology. If the model fit well to all the
different promoters, this would increase our confidence that it would serve well
as a foundation for inferring repressor kinetics later in
Section~\ref{sec:rep_kinetics_inference}. Conversely, were the model to fail on
more than a couple of the other promoters, it would give us pause.

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to $\ln k_i/\gamma \sim \text{constant} -
\beta\Delta\varepsilon_P$, described in text. Each individual promoter is
labeled with a unique number in both (C) and (D) for cross comparison.}
\label{fig:constit_post_full}
\end{figure}

\fig{fig:constit_post_full}(C) shows the results, plotting the posterior
distribution from individually MCMC sampling all 18 constitutive promoter
datasets from~\cite{Jones2014}. To aid visualization, rather than plotting
samples for each promoter's posterior as in \fig{fig:constit_post_full}(A), for
each posterior we find and plot the curve that surrounds the 95\% highest
probability density region. What this means is that each contour 
encloses approximately 95\% of the samples, and thus 95\% of the probability
mass, of its posterior distribution. Posterior predictive ECDFs (not shown)
display a similar level of agreement between data and predictive samples as for
the bursty model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
\mmnote{Might be worth showing some/all of these PPCs in supplement for
completeness? We could do postage stamp sized diff plots and probably fit all 18
promoters on 1-2 pages.}\mrm{Definitely need to show them in SI.}

One interesting feature from \fig{fig:constit_post_full}(C) is that burst rate
varies far more widely, over a range of $\sim10^2$, than burst size, confined to
a range of $\lesssim10^1$ (and with the exception of promoter 6, just a span of
3-5x). This suggests that $k_i$, not $b$, is the key dynamic variable that
promoter sequence tunes.

It is interesting to connect this observation to the work
of~\cite{Brewster2012}, where these same 18 promoters were considered through
the lens of the three-state equilibrium model (model 2 in
Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
previously discussed the thermodynamic models of gene regulation can only make
statements about the mean gene expression. This implies that we can draw the
connection between both frameworks by equating the mean mRNA $\left\langle m \right\rangle$. This results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma}
        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
\end{equation}
By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
\exp(-\beta\Delta\varepsilon_r \ll 1)$) results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
\end{equation}
valid for all the binding energies considered here.

Given this result, how are the two coarse-grainings related? The only
association that makes dimensional sense and produces the correct
order-of-magnitude for the known parameters is to take
\begin{equation}
\frac{k_i}{\gamma} = \exp(-\beta\Delta\varepsilon_P)
\label{eq:bursty_equil_corresp1}
\end{equation}
and
\begin{equation}
b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
\label{eq:bursty_equil_corresp2}
\end{equation}
Figure~\ref{fig:constit_post_full}(D) shows that this linear scaling between
$\ln k_i$ and $-\beta\Delta\varepsilon_P$ is approximately true for all 18
constitutive promoters considered. The intercept is an uninteresting function of
the unknown model parameters, so we only consider the slope of the scaling in
\fig{fig:constit_post_full}(D).
\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}

While the associations represented by
Eq.~\ref{eq:bursty_equil_corresp1} and Eq.~\ref{eq:bursty_equil_corresp2}
appear to be born out by the scaling $\ln k_i \sim -\beta\Delta\varepsilon_P$,
we do not find the association of parameters encoded by
\eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
intuitive. If some deeper insight is waiting to be gleaned from it, we leave it
as an open problem.
